---
title: "Cleaning and Wrangling Data"
author: "Alex Phillips"
date: "2023-04-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Packages 
```{r load packages, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyr)
library(readr)
```


# Introduction
This data was downloaded from the Alaska Department of Fish & Game containing commercial catch data from 1878-1997. The data set and reference to the original source can be found at its [public archive](https://knb.ecoinformatics.org/#view/df35b.304.2). Downloaded on April 4 2023. 

## Read Data

```{r loading data, warning=FALSE, message=FALSE}
catch_original <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/df35b.302.1")

```


## Explore Data

```{r explore, eval=FALSE, echo=FALSE}

colnames(catch_original)
head(catch_original) # gives us the six first rows of data
summary(catch_original)
sort(unique(catch_original$Year))

unique(catch_original$Region)

```


# Clean and Reshape Data

## Selecting and Removing Columns with `select()`

```{r selecting}

catch_data <- catch_original %>% 
  select(Region, Year, Chinook, Sockeye, Coho, Pink, Chum)

head(catch_data)

```


## Check Column Types 
```{r checking}

glimpse(catch_data)

```
## Changing column content using `mutate()`

```{r changing data type}

catch_clean <- catch_data %>%
  mutate(Chinook = as.numeric(Chinook))

head(catch_clean)

```

## Checking NAs

```{r checking NAs}

i <- which(is.na(catch_clean$Chinook))
catch_data[i,] 

```

## Fixing the Chinook column
```{r fixing Chinook}

catch_clean <- catch_data %>% 
  mutate(Chinook = if_else(Chinook == "I",
                           "1", 
                           Chinook),
         Chinook = as.numeric(Chinook))

## Check 

catch_clean[i,]

```



## Reshape Data 
From wide to long use `pivot_long()`
```{r}

catch_long <- catch_clean %>% 
  pivot_longer(cols = -c(Region, Year),
               names_to = "species",
               values_to = "catch")

head(catch_long)

```

From long to wide use `pivot_wider()`

```{r}
catch_wide <- catch_long %>% 
  pivot_wider(names_from = species, 
              values_from = catch)

head(catch_wide)
```

## renaming columns

```{r}
catch_long <- catch_long %>%
  rename(catch_thousands = catch) %>%
  mutate(catch = catch_thousands * 1000)

head(catch_long)
```

## Add a column using `mutate()`

```{r}
catch_long <- catch_long %>% 
  mutate(catch = catch_thousands*1000)  %>% 
  select(-catch_thousands)

head(catch_long)
```
## Split Apply Combine Workflow
We are going to use `groupby()` function and `summarize`

```{r}

mean_region <- catch_long %>% 
  group_by(Region) %>% 
  summarize(catch_mean = mean(catch))

head(mean_region)

mean_reg_yr <- catch_long %>% 
  group_by(Region, Year) %>% 
  summarize(catch_mean = mean(catch))

head(mean_reg_yr)


n_regions <- catch_long %>% 
  group_by(Region) %>% 
  summarize(observation_number =n())

head(n_regions)

```

## filtering by rows with `filter()` 

```{r filtering}

sse_catch <- catch_long %>% 
  filter(Region == "SSE")

head(sse_catch)

```



