---
title: "ggplot2 practice"
author: "Alex Phillips"
date: "2023-04-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Packages 

```{r packages, warning=FALSE, message=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(leaflet)
library(DT)
library(scales)
```

## Load Data

```{r load data, warning=FALSE, message=FALSE}
escapement <- read_csv("https://knb.ecoinformatics.org/knb/d1/mn/v2/object/urn%3Auuid%3Af119a05b-bbe7-4aea-93c6-85434dcb1c5e")
```

## Explore Data

```{r exploring }

## Check out column names
colnames(escapement)

## Peak at each column and class
glimpse(escapement)

## From when to when
range(escape$sampleDate)

## How frequent?
head(escape$sampleDate)
tail(escape$sampleDate)

## Which species?
unique(escape$Species)

```

## Calculate escape

```{r wrangling}
annual_esc <- escapement %>%
    separate(sampleDate, c("Year", "Month", "Day"), sep = "-") %>%
    mutate(Year = as.numeric(Year)) %>%
    group_by(Species, SASAP.Region, Year) %>%
    summarize(escape = sum(DailyCount)) %>%
    filter(Species %in% c("Chinook", "Sockeye", "Chum", "Coho", "Pink"))

head(annual_esc)
```


## Plotting 

Plot total escapement by species 
Minimal inputs are ggplot function, mapping aesthetics (x and y), data, geometry of the plot 

```{r Data Mapping}

## Option 1 - data and mapping called in the ggplot() function
ggplot(data = annual_esc,
       aes(x = Species, y = escapement)) +
    geom_col()

## Option 2 - data called in ggplot function; mapping called in geom
ggplot(data = annual_esc) +
    geom_col(aes(x = Species, y = escapement))


## Option 3 - data and mapping called in geom
ggplot() +
    geom_col(data = annual_esc,
             aes(x = Species, y = escapement))

```

```{r plot types}

annual_esc %>%
    filter(Year == 1974,
           Species %in% c("Chum", "Pink")) %>%
    ggplot(aes(x = Species, y = escapement)) +
    geom_boxplot()

annual_esc %>%
    filter(Year == 1974,
           Species %in% c("Chum", "Pink")) %>%
    ggplot(aes(x = Species, y = escapement)) +
    geom_violin()

annual_esc %>%
    filter(Species  == "Sockeye",
           SASAP.Region == "Bristol Bay") %>%
    ggplot(aes(x = Year, y = escapement)) +
    geom_line() +
    geom_point()

```

```{r customizing plots}

ggplot(annual_esc,
       aes(x = Species, y = escapement)) +
    geom_col(fill = "blue")

ggplot(annual_esc,
       aes(x = Species, y = escapement,
           fill = SASAP.Region)) +
    geom_col()

```
## Multiple Plots

We want data from a subset of years 

```{r subset data}

annual_esc_2000 <- annual_esc %>% 
  filter(Year %in% c(2000:2016))

# quick check 
unique(annual_esc_2000$Year)
head(annual_esc_2000)

```
Plot with `facet_wrap()`

```{r plot with faceting}

ggplot(annual_esc_2000,
       aes(x= Year,
           y= escape,
           color= Species)) +
  geom_line()+
  geom_point()+
  facet_wrap(~SASAP.Region,
             scales = "free_y")

```

## Customizing ggplot themes

`theme_*()` function

```{r themes}

ggplot(annual_esc_2000,
       aes(x= Year,
           y= escape,
           color= Species)) +
  geom_line()+
  geom_point()+
  facet_wrap(~SASAP.Region,
             scales = "free_y") + 
  theme_minimal()+
  ylab("Escapement")+
  ggtitle("Annual Salmon Escapement by Region")+
  theme(plot.title = element_text(hjust = 0.5))

```
Removing legend title and moving to bottom of plot 

```{r legends}

ggplot(annual_esc_2000,
       aes(x= Year,
           y= escape,
           color= Species)) +
  geom_line()+
  geom_point()+
  facet_wrap(~SASAP.Region,
             scales = "free_y") + 
  theme_minimal()+
  ylab("Escapement")+
  ggtitle("Annual Salmon Escapement by Region")+
  theme(legend.position = "bottom",
          legend.title = element_blank())


```

## Saving themes 

```{r saving theme}

my_theme <- theme_bw()+
  theme(legend.position = "bottom",
        legend.title = element_blank())

```

Creating plot with same theme
 
```{r plot with same theme }

ggplot(annual_esc_2000,
       aes(x = Year,
           y = escape,
           color = Species)) +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = seq(2000, 2016, 2)) +
    facet_wrap( ~ SASAP.Region,
                scales = "free_y") +
    ylab("Escapement") +
    ggtitle("Annual Salmon Escapement by Region") +
    my_theme +
    theme(axis.text.x = element_text(angle = 45,
                                     vjust = 0.5))

```
Smarter ticks using `scales()` 

```{r scale fixing}

annual_region_plot <- ggplot(annual_esc_2000,
       aes(x = Year,
           y = escape,
           color = Species)) +
    geom_line() +
    geom_point() +
    scale_x_continuous(breaks = seq(2000, 2016, 2)) +
    scale_y_continuous(labels = comma)+
    facet_wrap( ~ SASAP.Region,
                scales = "free_y") +
    ylab("Escapement") +
    ggtitle("Annual Salmon Escapement by Region") +
    my_theme +
    theme(axis.text.x = element_text(angle = 45,
                                     vjust = 0.5))

annual_region_plot

```

Saving plots

```{r saving}

ggsave("figs/annual_escape_region.png", annual_region_plot, width = 12, height = 8, units = "in")

```

## Maps with `leaflet`  

```{r}

location <- escapement %>% 
  distinct(Location, Latitude, Longitude) %>% 
  drop_na()

```

Create map with leaflet 
```{r}

leaflet(location) %>% 
  addTiles() %>% 
  addMarkers(lng= ~Longitude, 
             lat = ~Latitude,
             popup = ~Location)

```

```{r}

leaflet(location) %>%
    addWMSTiles(
        "https://www.gebco.net/data_and_products/gebco_web_services/web_map_service/mapserv?request=getmap&service=wms&BBOX=-90,-180,90,360&crs=EPSG:4326&format=image/jpeg&layers=gebco_latest&width=1200&height=600&version=1.3.0",
        layers = 'GEBCO_LATEST',
        attribution = "Imagery reproduced from the GEBCO_2022 Grid, WMS 1.3.0 GetMap, www.gebco.net"
    ) %>%
    addCircleMarkers(
        lng = ~ Longitude,
        lat = ~ Latitude,
        popup = ~ Location,
        radius = 5,
        # set fill properties
        fillColor = "salmon",
        fillOpacity = 1,
        # set stroke properties
        stroke = T,
        weight = 0.5,
        color = "white",
        opacity = 1
    )

```


