---
title: "RMarkdown Analysis"
output: 
  html_document:
    toc: true
    toc_float: true
date: "2023-04-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The data set we are analyzing comes from the [Arctic Data Center](https://arcticdata.io/catalog/view/doi:10.18739/A25T3FZ8X) and contains biogeochemical data from CTD measurements in the south Arctic ocean. This data set includes diverse measurements of seawater chemistry along with supporting conductivity, temperature, and depth (CTD) data. Measurements were taken from sea ice platforms each April or May from 2000-2014. Downloaded on April 3, 2023. 

### Read Data

#### Set up
```{r}
library(readr)
```


```{r read data, warning=FALSE, message=FALSE}
bg_chem <- readr::read_csv("data/BGchem2008data.csv")

```

#### Data Exploration 

```{r data exploration, eval=FALSE, echo=FALSE}
## column names
colnames(bg_chem)

## structure of my data
str(bg_chem)

##summary stats
summary(bg_chem)

## open data in tab 
View(bg_chem)

## unique values in a column - gives the indivdual ones that 
unique(bg_chem$Date)

```


## Analysis

### Calculate summary statistics

```{r summaries}
nitrate <- mean(bg_chem$NO3)
nitrite <- mean(bg_chem$NO2)
amm <- mean(bg_chem$NH4)
phos <- mean(bg_chem$P)
```


### Calculate mean Redfield ratio

Redfield ratio is 16:1 N:P 

```{r ratio calc}

ratio <- (amm + nitrate + nitrite)/phos
```


### Plot Redfield ratio

```{r plot}
plot(bg_chem$P, bg_chem$NO3 + bg_chem$NO2 + bg_chem$NH4)
```


## Conclusions

The redfield ratio is `r round(ratio)`

